WORCS: A Workflow for Open Reproducible Code in Science

### 2020-11-03 paper/Manuscript_DS.Rmd
Some touches

WORCS provides a user-friendly and
lightweight workflow for open, reproducible research, that meets all
TOP-guidelines, except replication.
Nevertheless, there are clear directions for future developments.
Firstly, although the workflow is currently implemented only in `R`, it is 
conceptually relevant for users of other statistical programming languages, most
notably `Python` [@10.5555/1593511].
We welcome efforts to implement WORCS in other platforms.
Secondly, even when a project is in `R`, it may have dependencies outside of the
`R` environment that cannot be managed using `renv`.
It is beyond the scope of the `worcs` package to support tools outside of `R`,
or to containerize a project so that it can
be identically reinstated on a different system or virtual machine.
The forthcoming package `repro` [@repro] will offer such extensions of the
workflow, thus combining the strengths of WORCS with the solutions proposed
by @peikertReproducibleDataAnalysis2019.

>>>>

WORCS provides a user-friendly and
lightweight workflow for open, reproducible research, that meets all
TOP-guidelines, except replication.
Nevertheless, there are clear directions for future developments.
Firstly, although the workflow is currently implemented only in `R`, it is 
conceptually relevant for users of other statistical programming languages, and we welcome efforts to implement WORCS in other platforms.
We welcome efforts to implement WORCS in other platforms.
Secondly, even when a project is in `R`, it may have dependencies outside of the
`R` environment that cannot be managed using `renv`.
It is beyond the scope of the `worcs` package to support tools outside of `R`,
or to containerize a project so that it can
be identically reinstated on a different system or virtual machine.
The forthcoming package `repro` [@repro] will offer such extensions of the
workflow, thus combining the strengths of WORCS with the solutions proposed
by @peikertReproducibleDataAnalysis2019.

### 2020-11-03 paper/reviews_ds.Rmd
Some touches

\RC{C1) The process as described by the author for conducting research is assumed to be linear. However, researchers often go back and forth on steps that they have already completed, and sometimes have to undo them completely and start from scratch. This does not seem to be tackled by the tool.}

@@

To address the Reviewer's comment, we have made two changes: First, we now explicitly state, in the paper and the Workflow vignette, that the steps are not necessarily linear:  *"Note that, although the steps are numbered for reference purposes, we acknowledge that the process of conducting research is not always linear."*. Second, while it was always possible to manually add a manuscript or preregistration at a later point in time after creating a new project in RStudio - we have now added functions to increase the non-linear flexibility of the `worcs` package: `add_manuscript()` and `add_preregistration()`.

@@

\RC{C2) The gist of the tool seems to allow researchers to document their steps and generate a scholarly document that is ready for submission. In a sense, there is a similarity with existing notebooks such as Jupyter. So I was wondering why don’t the authors start from an existing popular tool such as Jupyter as the starting point (given the similarity in features that they would like to add), and extend Jupyter for instance with the new features such as dependency documentation and version control ?}

Response: There are two points we wish to address here. First, we would like to clarify that this paper introduces WORCS, a conceptual workflow for open science research projects. When the Reviewer refers to "the tool" - we assume this to mean the `worcs` R-package, which offers convenience functions to help R-users follow the WORCS-workflow. This is clarified in the Abstract and in the paper, e.g.:

> Abstract: *This paper introduces the Workflow for Open Reproducible Code in Science (WORCS): A step-by-step procedure that researchers can follow to make a research project open and reproducible, accompanied by an open source software implementation that automates these steps.*

> Manuscript, paragraph titled **The R implementation of WORCS**: *WORCS is, first and foremost, a conceptual workflow that could be implemented in any software environment. As of this writing, the workflow has been implemented for `R` users in the package `worcs` [@worcspackage2020].*

Second, in response to the main comment, we agree with the Reviewer that the conceptual workflow as explained in the paper could be implemented in any software environment, including Jupyter. In fact, we emphasize this in the paragraph on *Future developments*: *"although the workflow is currently implemented only in `R`, it is conceptually relevant for users of other statistical programming languages, most notably `Python` [@10.5555/1593511]. We welcome efforts to implement WORCS in other platforms."*.

@@

\RC{C3) Regarding Version control and dependency documentation, the authors uses in their system existing solution, namely git and env, respectively. While I did not criticize reusing existing solution, on the contrary, I don't believe that simply using such solution add a substantial value. For example, git capture changes at the level of the line of code/text, whereas a researcher needs an abstraction that informs him/her on the coarse-grained elements that are significant from the point of view of development and research steps so as to have a clear idea of where s/he is within the process and help him/her make the decision of the next steps to undertake.}

@@

Second, the Reviewer claims that using e.g. Git does not add substantial value - but this claim is in disagreement with published literature that makes a strong case for the value of using Git for research [see two references cited in the paper, @ramGitCanFacilitate2013;  @blischakQuickIntroductionVersion2016].

Third, the Reviewer remarks that 
"git capture changes at the level of the line of code/text, whereas a researcher needs an abstraction that informs him/her on the coarse-grained elements [...] so as to have a clear idea of where s/he is within the process and help him/her make the decision of the next steps to undertake".
This comment suggests a fundamental misunderstanding of how Git works.
Although Git tracks file changes on a line-by-line basis,
any number of such changes can be committed together, thereby capturing the more "course-grained elements".
For example, in our response to these reviews, we have created one Git commit for each Reviewer comment, and these commits contain changes to the action letter, manuscript, and vignettes.

We do see the value in clarifying this point in the manuscript as well, and to address this comment, the manuscript now reads:

>>>>

\RC{C1: The process as described by the author for conducting research is assumed to be linear. However, researchers often go back and forth on steps that they have already completed, and sometimes have to undo them completely and start from scratch. This does not seem to be tackled by the tool.}

@@

To address the Reviewer's comment, we have made two changes: First, we now explicitly state, in the paper and the Workflow vignette, that the steps are not necessarily linear:  *"Note that, although the steps are numbered for reference purposes, we acknowledge that the process of conducting research is not always linear."*. Second, while it was always possible to manually add a manuscript or preregistration at a later point in time after creating a new project in RStudio - we have added more functions to increase the non-linear flexibility of the `worcs` package: `add_manuscript()` and `add_preregistration()`.

@@

\RC{C2: The gist of the tool seems to allow researchers to document their steps and generate a scholarly document that is ready for submission. In a sense, there is a similarity with existing notebooks such as Jupyter. So I was wondering why don’t the authors start from an existing popular tool such as Jupyter as the starting point (given the similarity in features that they would like to add), and extend Jupyter for instance with the new features such as dependency documentation and version control ?}

Response: There are two points we wish to address here. First, the Reviewer focuses on "the tool" - which we assume refers to the R package. However, the main goal of this paper is to introduce the conceptual workflow. The R package is merely offered to make adoption of the conceptual workflow easier for R users. This distinction is clarified several places throughout the manuscript, and to prevent further confusion, we have clarified the distinction even further in the revision, e.g.:

Abstract:

> This paper introduces the Workflow for
  Open Reproducible Code in Science (WORCS): A step-by-step
  procedure that researchers can follow to make a research project open and
  reproducible.  
  [...]  
  To facilitate widespread adoption, the WORCS principles have been implemented
  in the R package `worcs`, which offers an RStudio 
  project template and utility functions for specific workflow steps.

Manuscript, introduction:

> This paper introduces the conceptual workflow (referred to in upper case, "WORCS"),
and discusses its underlying principles and how it meets different standards for open science.
The principles underlying this workflow are universal and largely platform independent.
We also present a software implementation of WORCS for `R` users [@rcoreteamLanguageEnvironmentStatistical2020]:
The `R` package `worcs` [referred to in monospace font, @worcspackage2020].

Manuscript, paragraph titled **The R implementation of WORCS**:

> WORCS is, first and foremost, a conceptual workflow that could be implemented in any software environment.
As of this writing, the workflow has been implemented for `R` users in the package `worcs` [@worcspackage2020].

Second, in response to the question "why don’t the authors start from an existing popular tool such as Jupyter",
<!--we are sure the Reviewer does not mean to imply that R is not a popular tool,
or that R users should switch to Jupyter if they wish to make their research open and reproducible.
To provide some context, on [ITJobsWatch](https://www.itjobswatch.co.uk),
R is the 216th highest ranked job requirement, and Jupyter is 648th.-->
we completely agree with the Reviewer that the conceptual workflow as explained in the paper could be implemented in any software environment, including Jupyter.
In fact, we emphasize this in several places throughout the manuscript, for instance the paragraph titled **The R Implementation of WORCS**:

And in **Future developments**:

> although the workflow is currently implemented only in `R`, it is conceptually relevant for users of other statistical programming languages, and we welcome efforts to implement WORCS in other platforms."

@@

\RC{C3: Regarding Version control and dependency documentation, the authors uses in their system existing solution, namely git and env, respectively. While I did not criticize reusing existing solution, on the contrary, I don't believe that simply using such solution add a substantial value. For example, git capture changes at the level of the line of code/text, whereas a researcher needs an abstraction that informs him/her on the coarse-grained elements that are significant from the point of view of development and research steps so as to have a clear idea of where s/he is within the process and help him/her make the decision of the next steps to undertake.}

@@

Second, the Reviewer claims that using e.g. Git does not add substantial value - but this claim is in disagreement with published literature (which we cite) that makes a strong case for the value of using Git for research [see @ramGitCanFacilitate2013;  @blischakQuickIntroductionVersion2016].

Third, there seems to be some misunderstanding of the workings of existing solutions.
For example, the Reviewer remarks that 
"git capture changes at the level of the line of code/text, whereas a researcher needs an abstraction that informs him/her on the coarse-grained elements [...] so as to have a clear idea of where s/he is within the process and help him/her make the decision of the next steps to undertake".
This comment suggests a fundamental misunderstanding of how Git works.
Although Git tracks file changes on a line-by-line basis,
any number of such changes can be committed together, thereby capturing the more "course-grained elements".
For example, in our response to these reviews, we have created one Git commit for each Reviewer comment, and these commits contain changes to the action letter, manuscript, and vignettes.

The Reviewer's comment does illustrate the importance of clarifying this point in the manuscript as well
To address this comment, the revised manuscript now reads:
